---
title: ".b[Descriptive Statistics, pt. I]"
subtitle: ".b[ECON 3640--001]"
author: "Marcio Santetti"
date: "Spring 2022"
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'utah-css.css']
    # self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      ratio: "16:9"
      countIncrementalSlides: false
---
class: inverse, middle

```{r Setup, include = F}
options(htmltools.dir.version = FALSE)
library(pacman)
p_load(broom, latex2exp, ggplot2, ggthemes, ggforce, viridis, dplyr, magrittr, knitr, parallel, xaringanExtra, tidyverse, sjPlot, showtext, mathjaxr, ggforce, furrr, kable, kableExtra, wooldridge, hrbrthemes, scales, ggeasy)




# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  warning = F,
  message = F,
  dpi=300
)

```



# Motivation

---


# Getting to know our data

<br><br><br>

Before any complex operations, we need to .hi[know our data].

--

Such previous assessment may .hi-blue[not] appear in a final work (e.g., a *report*, a *paper*, a *project*), but is the .hi[first step] of any data analysis procedure.




---

layout: false
class: inverse, middle

# Some key statistical concepts


---

# Key statistical concepts

<br>

Two of the *most important* statistical concepts:

--

- .hi[Population]: a group of *all* items of interest to a statistics practitioner.

  - A descriptive measure of a population is called a .red[*parameter*].
  

--

<br>

- .hi-blue[Sample]: A *set* of data drawn from the studied population.

  - A descriptive measure of a sample is called a .blue[*statistic*].
  
  - It is used to make *inferences* about population parameters.

---


layout: false
class: inverse, middle

# Types of data and information


---

# Types of data and information

<br><br>

In Statistics, a .hi[variable] is a characteristic of a population or sample.

--

<br>

3 main .hi-blue[types] of data in Statistics:

  - *Interval* data;
  
  - *Nominal* data;
  
  - *Ordinal* data.
  

---

# Types of data and information

<br><br>

.hi[Interval data ]is also known as *quantitative* or *numerical* data.
  
  - Examples?

--

.hi[Nominal data] comprehends *qualitative* or *categorical* variables.

  - Ordering *does not* matter.
  
  - Examples?
  
--

For .hi[ordinal data], ordering *does* matter.

  - Examples?


---

layout: false
class: inverse, middle

# Graphical statistical techniques


---

# Graphical statistical techniques

<br><br><br>

Sometimes, a *table* describing nominal (or ordinal) data may not be the best option to catch a reader’s eye.

--

Therefore, some graphical techniques can be useful, such as the .hi-blue[bar] and the .hi-blue[pie] charts.


---


# Graphical statistical techniques

A .hi-slate[bar chart] is useful to illustrate an .red[*absolute frequency*], i.e., the total number of observations fitting a given category in a data set.

--

```{r, echo=FALSE, warning=FALSE, message=FALSE, dev='svg', fig.height = 5}

font_add_google(name = "Roboto Condensed", family = "Roboto Condensed")

titles <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-04-20/netflix_titles.csv')


titles %>% 
  filter(type == "Movie") %>% 
  count(rating, sort=TRUE) %>%
  filter(! rating %in% c(NA, "TV-Y7-FV", "UR", "NC-17")) %>% 
  ggplot() + geom_col(aes(y = fct_reorder(rating, n), x = n), fill = "steelblue4") +
  theme_ipsum_rc() +
  labs(x = '# Movies', y = "Ratings",
       title = 'Netflix movies by category') +
  scale_x_comma() +
  easy_x_axis_title_size(13) +
  easy_y_axis_title_size(13)

```


---

# Graphical statistical techniques

If, on the other hand, we want to illustrate .red[*relative frequencies*], i.e., the percentage with which each category appears in the data set, a .hi-slate[pie chart] is the best option.

--

```{r, echo=FALSE, warning=FALSE, message=FALSE, dev='svg', fig.height = 5}

titles %>% 
  count(type, sort = TRUE) %>% 
  mutate(share = round(n/sum(n)*100, 2)) %>% 
  ggplot(aes(x="", y = n, fill = type)) + 
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  theme_void() +
  easy_remove_x_axis() +
  easy_remove_y_axis() +
  labs(title = "Netflix catalog by type") +
  theme(text = element_text(family = "Roboto Condensed")) +
  geom_text(aes(label = paste0(share, "%")), position = position_stack(vjust=0.5),
            color=c("white", "black")) +
  viridis::scale_fill_viridis(discrete = TRUE, option = "D") +
  easy_add_legend_title("Movie or TV Show?") +
  easy_plot_legend_size(11) +
  easy_plot_title_size(17)

```
---

# Graphical statistical techniques

<br>

Moving on to .hi[interval data], one of the most common ways of presenting it is through a .hi-slate[histogram].

--

<br>

To construct a histogram, the easiest way to start is by following a .red[*recipe*]:

1. Find the data set’s *lowest* and the *highest* values;

2. Define the appropriate intervals (*bin size*), and the number of *observations* contained in each interval;

3. Draw *each* bin next to each other.




---

# Graphical statistical techniques

An example:

```{r, echo=FALSE, warning=FALSE, message=FALSE, dev='svg', fig.height = 5}


titles %>% 
  filter(type == "Movie") %>% 
  separate(duration, c("duration_mins", "B")) %>% 
  mutate(duration_mins = as.integer(duration_mins)) %>% 
  ggplot() + 
  geom_histogram(aes(duration_mins), bins = 12, color = "white", fill = "lightcoral") +
  theme_ipsum_rc() +
  labs(title = "Netflix movie lengths in minutes", y = "Count", x = "Minutes") +
  easy_x_axis_title_size(12) +
  easy_y_axis_title_size(12)


```


---

# Graphical statistical techniques

<br><br><br>

Suppose that we want to visually describe the relationship between .hi-blue[two] interval variables.

--

The technique to be used is the .hi-slate[scatter diagram].

--

To draw a scatter plot, we must define an .red[*independent*] (explanatory) and a .red[*dependent*] (explained) variable.

---

# Graphical statistical techniques

Example 1:

```{r, echo=FALSE, warning=FALSE, message=FALSE, dev='svg', fig.height = 5.5}

data("sleep75")

sleep75 %>% as_tibble() %>% 
  ggplot(aes(y = sleep, x = worknrm)) +
  geom_point(size = 2, color = "brown1", alpha = 0.7) +
  labs(x = "Minutes worked per week", y = "Minutes slept per week",
       title = "Sleeping vs. working time") +
  theme_ipsum_rc() +
  easy_x_axis_title_size(12) +
  easy_y_axis_title_size(12)

```


---

# Graphical statistical techniques

Example 2:


```{r, echo=FALSE, warning=FALSE, message=FALSE, dev='svg', fig.height = 5.5}

data("wage2")

wage2 %>% as_tibble() %>% 
  ggplot(aes(y = wage, x = hours)) +
  geom_point(size = 2, color = "brown1", alpha = 0.7) +
  labs(x = "Hours worked per week", y = "Monthy earnings",
       title = "Wage vs. hours worked") +
  theme_ipsum_rc() +
  easy_x_axis_title_size(12) +
  easy_y_axis_title_size(12)



```

---



layout: false
class: inverse, middle

# Next time: Statistical measures


---
exclude: true