\documentclass[11pt,]{article}
\usepackage[margin=1in]{geometry}
\newcommand*{\authorfont}{\fontfamily{phv}\selectfont}
\usepackage[]{mathpazo}
\usepackage{abstract}
\renewcommand{\abstractname}{}    % clear the title
\renewcommand{\absnamepos}{empty} % originally center
\newcommand{\blankline}{\quad\pagebreak[2]}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\usepackage{longtable,booktabs}

\usepackage{parskip}
\usepackage{titlesec}
\titlespacing\section{0pt}{12pt plus 4pt minus 2pt}{6pt plus 2pt minus 2pt}
\titlespacing\subsection{0pt}{12pt plus 4pt minus 2pt}{6pt plus 2pt minus 2pt}

\titleformat*{\subsubsection}{\normalsize\itshape}

\usepackage{titling}
\setlength{\droptitle}{-.25cm}

%\setlength{\parindent}{0pt}
%\setlength{\parskip}{6pt plus 2pt minus 1pt}
%\setlength{\emergencystretch}{3em}  % prevent overfull lines

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}

\usepackage{fancyhdr}
\pagestyle{fancy}
\usepackage{lastpage}
\renewcommand{\headrulewidth}{0.3pt}
\renewcommand{\footrulewidth}{0.0pt}
\lhead{}
\chead{}
\rhead{\footnotesize \textbf{ECON 3640--001: PROBABILITY \& STATISTICAL
INFERENCE} -- Spring 2022}
\lfoot{}
\cfoot{\small \thepage/\pageref*{LastPage}}
\rfoot{}

\fancypagestyle{firststyle}
{
\renewcommand{\headrulewidth}{0pt}%
   \fancyhf{}
   \fancyfoot[C]{\small \thepage/\pageref*{LastPage}}
}

%\def\labelitemi{--}
%\usepackage{enumitem}
%\setitemize[0]{leftmargin=25pt}
%\setenumerate[0]{leftmargin=25pt}




\makeatletter
\@ifpackageloaded{hyperref}{}{%
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
}
\@ifpackageloaded{color}{
    \PassOptionsToPackage{usenames,dvipsnames}{color}
}{%
    \usepackage[usenames,dvipsnames]{color}
}
\makeatother
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
             pdfkeywords = {},
            pdftitle={\textbf{ECON 3640--001: PROBABILITY \& STATISTICAL
INFERENCE}},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=blue,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls


\setcounter{secnumdepth}{0}

\usepackage{longtable}




\usepackage{setspace}

\title{\textbf{ECON 3640--001: PROBABILITY \& STATISTICAL INFERENCE}}
\date{Spring 2022}


\begin{document}

		\maketitle
	

		\thispagestyle{firststyle}

%	\thispagestyle{empty}


  	\noindent \begin{tabular*}{\textwidth}{ @{\extracolsep{\fill}} lr @{\extracolsep{\fill}}}

  \hline
  \\



\textbf{Instructor}: Marcio
Santetti & \textbf{E-mail}: \texttt{\href{mailto:marcio.santetti@utah.edu}{\nolinkurl{marcio.santetti@utah.edu}}}\\
\textbf{Office Hours}: Before or after class, or by
appointment  &  \textbf{Time}: TuTh, 2:00PM--3:20PM\\
\textbf{Web}: \href{http://utah.instructure.com/courses/767112}{\tt utah.instructure.com/courses/767112}  & \textbf{Place}: AEB
330\\
	&  \\
	\hline
	\end{tabular*}

\vspace{2mm}



\hypertarget{overview}{%
\section{Overview}\label{overview}}

\textbf{\emph{Important}}: For guidelines and practical recommendations
concerning COVID-19, please constantly check out
\href{https://coronavirus.utah.edu/}{\texttt{https://coronavirus.utah.edu/}}.

This is an applied course. Beyond introducing students to the most
important descriptive, probabilistic, and inferential tools from both
classical (frequentist) and Bayesian approaches, everyone will have the
opportunity to apply these concepts to several different data sets and
research questions.

\vspace{.2cm}

\hypertarget{main-references}{%
\section{Main References}\label{main-references}}

In addition to lecture notes, that come from several references, other
readings will be posted on Canvas.

\bigskip

\textbf{Recommended textbooks}:

\begin{itemize}
\item
  G. Keller, \emph{Statistics for Management and Economics}, 10th
  edition, South-Western/Cengage, 2014.
\item
  A. Johnson, M. Ott, and M. Dogucu. \emph{Bayes Rules! An Introduction
  to Applied Bayesian Modeling}, CRC Press, 2022.

  \begin{itemize}
  \tightlist
  \item
    Available online at
    \href{https://www.bayesrulesbook.com/}{\texttt{https://www.bayesrulesbook.com/}}.
  \end{itemize}
\end{itemize}

\bigskip

\textbf{Additional References} (optional):

\begin{itemize}
\item
  J. Blitzstein and J. Hwang. \emph{Introduction to Probability},
  2\(^{nd}\) edition, Chapman \& Hall/CRC Texts in Statistical Science,
  2019.
\item
  J. Albert and J. Hu. \emph{Probability and Bayesian Modeling}, CRC
  Press, 2020.
\end{itemize}

\bigskip

\hypertarget{software-requirements}{%
\section{Software Requirements}\label{software-requirements}}

\vspace{.3cm}

Our classes will always have \emph{theoretical} and \emph{applied}
components. The latter will be done using \(\textsf{R}\), an open-source
and completely \emph{free} statistical/programming language. In applied
lectures, I use the \(\textsf{R}\)Studio IDE (Integrated Development
Environment), also freely available. However, students are welcome to
use the platform they prefer.

In addition, having a spreadsheet editor (e.g., Excel) is strongly
recommended. University of Utah students should have free access to
Microsoft Office programs through
\href{https://o365cloud.utah.edu/}{\texttt{Office\ 365}}.

No previous knowlegde in \(\textsf{R}\) is required. The course will
guide students through every step in the applied lectures.

\begin{itemize}
\tightlist
\item
  Download \(\textsf{R}\):
  \href{https://cloud.r-project.org/}{\texttt{https://cloud.r-project.org/}}
\item
  Download \(\textsf{R}\)Studio:
  \href{https://rstudio.com/products/rstudio/}{\texttt{https://rstudio.com/products/rstudio/}}
\end{itemize}

\bigskip

\hypertarget{prerequisites}{%
\section{Prerequisites}\label{prerequisites}}

\vspace{.1cm}

College algebra, ECON 2010 and ECON 2020.

\hypertarget{course-outcomes}{%
\section{Course Outcomes}\label{course-outcomes}}

At the completion of this course, you will be able to:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Comfortably use \(\textsf{R}\) to perform basic statistical analyses;
\item
  Manipulate and extract relevant information from diverse data sets;
\item
  Perform statistical inference from both classical and Bayesian
  approaches;
\item
  Critically assess the most common tools in Statistics.
\end{enumerate}

\hypertarget{assignments-exams-and-grading}{%
\section{Assignments, Exams, and
Grading}\label{assignments-exams-and-grading}}

\begin{itemize}
\item
  \textbf{Class participation}: In the absence of symptoms and contact
  with potentially infected individuals, attendance is a crucial part of
  the course.
\item
  \textbf{Problem Sets}: Whenever enough content is seen in class, a new
  problem set will be uploaded to our Canvas course page. Students are
  \emph{not} required to turn these in, so they can take their own time
  to complete them. The instructor will upload the answer keys two days
  before each exam. Thus, do not rely on waiting for the answer sheets
  to prepare for exams.
\item
  \textbf{Applied research project} (30\%): Since this is an applied
  course, students will be asked to select a data set of their interest
  and perform statistical analyses on it. Projects may involve thorough
  descriptive work, inferential analyses, etc. We will have several
  checkpoints throughout the semester to track everyone's progress. The
  project must be between 8 and 10 pages, double-spaced. All necessary
  assistance will be provided by the instructor upon students' requests.
\item
  \textbf{Midterm exam} (30\%): The midterm exam will ask theoretical
  and applied questions on descriptive statistics and some probability
  concepts.
\item
  \textbf{Final exam} (40\%): The Final exam is comprehensive, covering
  all studied topics. In principle, the final exam will also be
  take-home. As we get closer, new procedures may be adopted by the
  University. We will discuss an appropriate window allowing every
  student to have time to complete the exams.
\end{itemize}

\textbf{\emph{Important note 1}}: Exams will strongly reflect the
content studies and practiced in class.

\textbf{\emph{Important note 2}}: For \emph{any} assignment, if an
answer contains a direct ``Copy+Paste'' from the lecture notes, or if it
is copied from someone else's work, \textbf{no credit will be given}.
Read more in the \emph{Academic Dishonesty} section, on page 5.

\textbf{\emph{Important note 3}}: Throughout the semester, several
opportunities for \textbf{extra credit} will be made available both in
class and on Canvas.

\hypertarget{important-dates}{%
\section{Important Dates}\label{important-dates}}

\vspace{.3cm}

\begin{itemize}
\tightlist
\item
  \textbf{Classes begin}: Jan 10
\item
  \textbf{Last day to add/drop classes}: Jan 21
\item
  \textbf{Martin Luther King Jr.~Day holiday}: Jan 17
\item
  \textbf{Presidents Day holiday}: Feb 21
\item
  \textbf{Midterm Exam}: Feb 10
\item
  \textbf{Last day to withdraw from classes}: Mar 4
\item
  \textbf{Spring Break}:Mar 6--13
\item
  \textbf{Final Exam}: May 3, time TBA
\end{itemize}

\hypertarget{letter-grade-distribution}{%
\section{Letter Grade Distribution}\label{letter-grade-distribution}}

\vspace{.2cm}

\begin{longtable}[]{@{}lc@{}}
\toprule
\textbf{\emph{Performance}} & \textbf{\emph{Letter Grade}} \\
\midrule
\endhead
Excellent, superior performance & A (93-100\%), A- (90-92.9\%) \\
Good performance & B+ (87-89.9\%), B (83-86.9\%), B- (80-82.9\%) \\
Standard performance & C+ (77-79.9\%), C (73-76.9\%), C- (70-72.9\%) \\
Substandard performance & D+ (67-69.9\%), D (63-66.9\%), D-
(60-62.9\%) \\
Unsatisfactory performance & E (0-59.9\%) \\
\bottomrule
\end{longtable}

\vspace{.5cm}

\hypertarget{class-policies}{%
\section{Class Policies}\label{class-policies}}

\vspace{.3cm}

You can expect me to:

\begin{itemize}
\item
  Ask exam questions reflecting what has been studied in class and
  viewed in questions from Problem Sets;
\item
  Grade and provide feedback on assignments and exams within two weeks
  from the due date;
\item
  Reply to emails/messages within 24 hours during the week and within 48
  hours on weekends and holidays;
\item
  Provide time for office hours as stated on page 1, where students can
  ask any question and talk about any issues/concerns relative to our
  course;
\item
  Ignore emails asking me to raise grades by the end of the semester.
\end{itemize}

I expect students to:

\begin{itemize}
\item
  Come to class in time;
\item
  Check out Announcements and new content updates as soon as possible;
\item
  Take the exams on the scheduled dates. No make-up exams will be
  allowed;
\item
  Respectfully participate in in-class discussions and activities;
\item
  Immediately notify me in the event of an emergency that prevents you
  from doing an exam or completing the course;
\item
  Ask questions if any expectations or assignments are unclear.
\end{itemize}

\hypertarget{tentative-course-schedule}{%
\section{Tentative Course Schedule}\label{tentative-course-schedule}}

\vspace{.3cm}

\begin{itemize}
\item
  \textbf{Week 1}: Course overview and introduction.
\item
  \textbf{Week 2}: Descriptive Statistics I.
\item
  \textbf{Week 3}: Descriptive Statistics II.
\item
  \textbf{Week 4}: Probability Theory I.
\item
  \textbf{Week 5}: Probability Theory II.
\item
  \textbf{Week 6}: Bayesian Estimation I.
\item
  \textbf{Week 7}: Bayesian Estimation II.
\item
  \textbf{Week 8}: Frequentist Statistics I.
\item
  \textbf{Week 9}: Frequentist Statistics II.
\item
  \textbf{Week 10}: Hypothesis Testing I.
\item
  \textbf{Week 11}: Hypothesis Testing II.
\item
  \textbf{Week 12}: Case Studies I.
\item
  \textbf{Week 13}: Case Studies II.
\item
  \textbf{Week 14}: Open for adjustments.\footnote{This week is open for
    adjustments, in case we need to review some week's content or
    assignment.}
\item
  \textbf{Week 15}: Course wrap-up and review.
\end{itemize}

\vspace{1cm}

\hypertarget{important-note}{%
\section{Important Note}\label{important-note}}

\vspace{.3cm}

This syllabus is meant to serve as an outline and guide for our course.
Please note that I may modify it with reasonable notice to you. I may
also modify the Course Schedule to accommodate the needs of our class.
Any changes will be announced in class and posted on Canvas under
\emph{Announcements}.

\hypertarget{institutional-policies-and-procedures}{%
\section{Institutional Policies and
Procedures}\label{institutional-policies-and-procedures}}

\vspace{.2cm}

\hypertarget{faculty-and-student-responsibilities}{%
\subsection{Faculty and Student
Responsibilities}\label{faculty-and-student-responsibilities}}

All students are expected to maintain professional behavior in the
classroom setting, according to the Student Code, spelled out in the
Student Handbook. Students have specific rights in the classroom as
detailed in Article III of the Code. The Code also specifies proscribed
conduct (Article XI) that involves cheating on tests, plagiarism, and/or
collusion, as well as fraud, theft, etc. Students should read the Code
carefully and know they are responsible for the content. According to
Faculty Rules and Regulations, it is the faculty responsibility to
enforce responsible classroom behaviors, beginning with verbal warnings
and progressing to dismissal from class and failing grade. Students have
the right to appeal such action to the Student Behavior Committee.
``Faculty . . . must strive in the classroom to maintain a climate
conducive to thinking and learn- ing.'' PPM 8-12.3, B.

``Students have a right to support and assistance from the University in
maintaining a climate conducive to thinking and learning.'' PPM 8-10,
II. A.

\hypertarget{wellness-statement}{%
\subsection{Wellness Statement}\label{wellness-statement}}

Personal concerns such as stress, anxiety, relationship difficulties,
depression, cross-cultural differences, etc., can interfere with a
students ability to succeed and thrive at the University of Utah. For
helpful resources contact the Center for Student Wellness:
\href{www.wellness.utah.edu}{\texttt{www.wellness.utah.edu}},
801-581-7776.

\hypertarget{student-names-personal-pronouns}{%
\subsection{Student Names \& Personal
Pronouns}\label{student-names-personal-pronouns}}

Class rosters are provided to the instructor with the students legal
name as well as ``Preferred first name'' (if previously entered by you
in the student profile section of your CIS account). While CIS refers to
this as merely a preference, I will honor you by referring to you with
the name and pronoun that feels best for you in class, on papers, exams,
group projects, etc. Please advise me of any name or pronoun changes
(and update CIS) so I can help create a learning environment in which
you, your name, and your pronoun will be respected.

\hypertarget{academic-dishonesty}{%
\subsection{Academic Dishonesty}\label{academic-dishonesty}}

The instructor of this course will take appropriate actions in response
to Academic Dishonesty, as defined the Universitys Student Code. Acts of
academic dishonesty include but are not limited to:

\begin{itemize}
\item
  Cheating: using, attempting to use, or providing others with any
  unauthorized assistance in taking quizzes, tests, examinations, or in
  any other academic exercise or activity. Unauthorized assistance
  includes:

  \begin{itemize}
  \tightlist
  \item
    Working in a group when the instructor has designated that the quiz,
    test, examination, or any other academic exercise or activity be
    done individually;
  \item
    Depending on the aid of sources beyond those authorized by the
    instructor in writing papers, preparing reports, solving problems,
    or carrying out other assignments;
  \item
    Substituting for another student, or permitting another student to
    substitute for oneself, in taking an examination or preparing
    academic work;
  \item
    Acquiring tests or other academic material belonging to a faculty
    member, staff member, or another student without express permission;
  \item
    Continuing to write after time has been called on a quiz, test,
    examination, or any other academic exercise or activity;
  \item
    Submitting substantially the same work for credit in more than one
    class, except with prior approval of the instructor; or engaging in
    any form of research fraud.
  \end{itemize}
\item
  Falsification: altering or fabricating any information or citation in
  an academic exercise or activity.
\item
  Plagiarism: representing, by paraphrase or direct quotation, the
  published or unpublished work of another person as ones own in any
  academic exercise or activity without full and clear acknowledgment.
  It also includes using materials prepared by another person or by an
  agency engaged in the sale of term papers or other academic materials.
\end{itemize}

\hypertarget{sexual-harassment}{%
\subsection{Sexual Harassment}\label{sexual-harassment}}

Title IX of the Education Amendments of 1972 prohibits sex
discrimination against any participant in an educational program or
activity that receives federal funds. The act is intended to eliminate
sex discrimination in education. Title IX covers discrimination in
programs, admissions, activities, and student-to-student sexual
harassment. It is the policy of the University of Utah to maintain an
academic and work environment free of sexual harassment for students,
faculty, staff and participants. A claim under this policy may be
brought by any faculty, staff member, student or participant based on
the conduct of any University employee or student that is related to or
in the course of University business. A claim may also be brought by an
administrator acting on behalf of the University. The Office of Equal
Opportunity and Affirmative Action (OEO/AA) will handle all alleged
sexual harassment matters pursuant to the Procedures set forth in Policy
5-210.

Any student, staff, faculty member, or participant in university
services or activities who believes that there is or has been sexual
harassment should contact any of the following:

\begin{itemize}
\tightlist
\item
  the cognizant academic chair of the department or the dean of the
  college within which the conduct occurred;

  \begin{itemize}
  \tightlist
  \item
    the immediate supervisor or director of the operational unit within
    which the conduct occurred;
  \item
    the Human Resources Division (Address: 420 Wakara Way, Suite 105,
    Salt Lake City, Utah 84108; Phone: 801-581-2169; TDD: 801-585-9070);
  \item
    directors or deans of Student Affairs and Services (Associate Dean -
    SW 108; Phone: 801-581- 8828);
  \item
    or the Office of Equal Opportunity and Affirmative Action (201 South
    Presidents Circle, John Park Building, Room 135, Salt Lake City,
    Utah 84112; Phone/TDD: 801-581-8365).
  \end{itemize}
\end{itemize}

Once informed, the supervisor, chair or dean shall consult with OEO/AA.
Similarly, OEO/AA shall advise as appropriate the cognizant dean, staff
director or administrator of any complaints of sexual harassment OEO/AA
receives concerning a member of the faculty, staff or student body.

\hypertarget{withdrawal-policy-and-i-grade}{%
\subsection{Withdrawal Policy and ``I''
Grade}\label{withdrawal-policy-and-i-grade}}

Failure to withdraw from school results in a E or EU grade being
recorded in all classes. Students may officially withdraw (\emph{W})
from a class or all classes after the drop deadline through the midpoint
of a course. A \emph{W} grade is recorded on the transcript and
appropriate tuition/fees are assessed. The grade of \emph{W} is not used
in calculating the students GPA.

An Incomplete grade can be given for work not completed due to
circumstances beyond your control. You must be passing the course and
have completed at least 80\% of the required coursework. Arrangements
must be made between you and the instructor concerning the completion of
the work. You may not retake a course without paying tuition. If you
attend class during a subsequent term, in an effort to complete the
coursework, you must register for the course. Once the work has been
completed, the instructor submits the grade to the Registrars Office.
The \emph{I} grade will change to an \emph{E} if a new grade is not
reported within one year. A written agreement between you and the
instructor may specify the grade to be given if the work is not
completed within one year. Copies of the agreement are kept by the
instructor and the academic department.

\hypertarget{americans-with-disabilities-act-ada-statement}{%
\subsection{Americans with Disabilities Act (ADA)
Statement}\label{americans-with-disabilities-act-ada-statement}}

The University of Utah seeks to provide equal access to its programs,
services and activities for people with disabilities. If you will need
accommodations in the class, reasonable prior notice needs to be given
to the Center for Disability Services, 162 Olpin Union Building,
581-5020 (V/TDD). CDS will work with you and the instructor to make
arrangements for accommodations. All written information in this course
can be made available in alternative format with prior notification to
the Center for Disability Services.

Additionally, the University endeavors to provide reasonable
accommodations and to ensure equal access to qualified persons with
disabilities. Inquiries concerning perceived discrimination or requests
for disability accommodations may be referred to the University's Title
IX/ADA/Section 504 Coordinator: Director, Office of Equal Opportunity
and Affirmative Action, 201 South Presidents Circle, Rm. 135, Salt Lake
City, UT, 84112. 801-581-8365 (voice/tdd), 801-585-5746 (fax).
\href{www.oeo.utah.edu}{\texttt{www.oeo.utah.edu}}.




\end{document}

\makeatletter
\def\@maketitle{%
  \newpage
%  \null
%  \vskip 2em%
%  \begin{center}%
  \let \footnote \thanks
    {\fontsize{18}{20}\selectfont\raggedright  \setlength{\parindent}{0pt} \@title \par}%
}
%\fi
\makeatother
