---
title: ".b[What is this course about?]"
subtitle: ".b[ECON 3640--001]"
author: "Marcio Santetti"
date: "Spring 2022"
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'utah-css.css']
    # self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      ratio: "16:9"
      countIncrementalSlides: false
---
class: inverse, middle

```{r Setup, include = F}
options(htmltools.dir.version = FALSE)
library(pacman)
p_load(broom, latex2exp, ggplot2, ggthemes, ggforce, viridis, dplyr, magrittr, knitr, parallel, xaringanExtra, tidyverse, sjPlot, showtext, mathjaxr, ggforce, furrr, kable, kableExtra,
       wooldridge, hrbrthemes, scales)




# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 5.5,
  fig.width = 10.5,
  warning = F,
  message = F,
  dpi=300
)

```



# Defining Statistics

---

# Trigger warning

<br><br><br>

.pull-left[

Learning this course's material may .hi[irreversibly] change the way you approach your own interests, and this fact may make it impossible to maintain your *current beliefs*.

]

.pull-right[

The learning journey may be very *stressful*, especially if you are not used to being .red[*frustrated].* 
]



---


# Defining Statistics

<br><br>

Statistics embodies the practice of .hi[collecting] and .hi[analyzing] data, either *quantitative* or *qualitative*.


<br>

--

Its purpose is to .hi-blue[describe], .hi-blue[measure], and .hi-blue[infer] relevant information from any data one has available.

--


<br><br><br>


.right[

But data are .note[fuzzy].
]



---


# Data are fuzzy

Some data from [`Mroz (1987)`](https://www.jstor.org/stable/1911029?seq=1#metadata_info_tab_contents):

.center[

```{r, echo=FALSE}
data(mroz)

mroz %>% as_tibble() %>% 
  select(hours, educ, wage, inlf, exper, motheduc, fatheduc, huswage, husage) %>% 
  head(15)
```

]

---

# Data are fuzzy

Some data from [`Mroz (1987)`](https://www.jstor.org/stable/1911029?seq=1#metadata_info_tab_contents):


```{r, echo=FALSE, warning=FALSE, message=FALSE, dev='svg', fig.height = 5}

font_add_google(name = "Roboto Condensed", family = "Roboto Condensed")

showtext_auto()

mroz %>% as_tibble() %>% 
  ggplot(aes(y = wage, x = educ)) +
  geom_point(color = '#4666d0', alpha=0.7, size=1.5) +
  labs(x = 'Years of education', y = 'Hourly wages') +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme_ipsum_rc()
```


---

# Data are fuzzy

```{r, echo=FALSE, warning=FALSE, message=FALSE, dev='svg'}



grades <- read_csv('sofar_grades.csv')


grades %>% 
  ggplot(aes(final_grade)) + 
  geom_bar(color='black', fill='#8b251c') +
  labs(x='', y='') +
  theme_ipsum_rc()

```

---

# Data are fuzzy

```{r, echo=FALSE, warning=FALSE, message=FALSE, dev='svg'}

showtext_auto()

grades <- read_csv('sofar_grades.csv')


grades %>% 
  ggplot(aes(final_grade)) + 
  geom_bar(color='black', fill='#8b251c') +
  labs(x="Final Letter Grades", y = "# Students", title = "Letter Grades for ECON 3640") +
  theme_ipsum_rc()

```


---

# Data are fuzzy

```{r, echo=FALSE, warning=FALSE, message=FALSE, dev='svg'}

data <- tibble(
  var1 = rnorm(1000, mean = 2, sd = 3),
  var2 = rnorm(1000, mean = 6, sd = 3),
  var3 = rnorm(1000, mean = 6, sd = 7)
)

data %>% ggplot() +
  geom_density(aes(x = var1, y = ..density..), fill = "#67dfc4", alpha = 0.5) +
  geom_density(aes(x = var2, y = ..density..), fill = "#b6134a", alpha = 0.6) +
  geom_density(aes(x = var3, y = ..density..), fill = "#9fb5b7", alpha = 0.6) +
  labs(x = '') +
  theme_ipsum_rc()
  


```


---

layout: false
class: inverse, middle

# Some practical tips


---

# Is this class hard?

<br>

- What does an Economist always have as a .hi[standard answer]?

--

  - It .hi-blue[depends]!
  
--

<br>

- Learning Statistics is quite a .red[journey]

--

  - But well worth it!
  - Will not be exhausted with this course
  - Here, a nicer challenge (theory and practice)
  - The instructor is here to _.red[help]_ you


---

# Is this class hard?

<br>

- In summary:

--

  - Take this class _.red[seriously]_
  - Come to class .red[in time] and with an .red[open mind]
  - Ask .red[questions]
  - Do the .red[assignments]
  - .red[Exams] will reflect what assignments (esp. _Problem Sets_) have asked you
  
--

- About .mono[R]

--

  - It may take some .red[time] to feel comfortable with it
  - But after a few weeks, it will become a .red[good friend]
  - Feed it well and it will be fine!
  
---

# A friend's advice

<br>

- Create a .hi[folder] in your computer for this course.

--
  
  - .hi-slate[Please].
  
--

<br>


- We will use .hi[hundreds] of different files throughout the semester.

  - .hi-slate[Organization] is key!
  - Even better: create folders for .hi-slate[each week]. It will make your life easier.
  


---



layout: false
class: inverse, middle

# Next time: Descriptive Statistics


---
exclude: true