---
title: ".b[Random Variables, pt. II]"
subtitle: ".b[ECON 3640--001]"
author: "Marcio Santetti"
date: "Spring 2022"
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'utah-css.css']
    # self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      ratio: "16:9"
      countIncrementalSlides: false
---
class: inverse, middle

```{r Setup, include = F}
options(htmltools.dir.version = FALSE)
library(pacman)
p_load(broom, latex2exp, ggplot2, ggthemes, ggforce, viridis, dplyr, magrittr, knitr, parallel, xaringanExtra, tidyverse, sjPlot, showtext, mathjaxr, ggforce, furrr, kableExtra, wooldridge, hrbrthemes, scales, ggeasy, patchwork)




# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  warning = F,
  message = F,
  dpi=300
)

theme_set(theme_ipsum_rc())

```



# Motivation



---

# Housekeeping

<br><br>


Notes based on `Keller (2009)`, ch. 8

---

# From discrete to continuous

Last week, we were introduced to .hi[random variables].

--

The starting point was to study .hi-blue[discrete] outcomes

  - That is, events from experiments that can be .hi-slate[listed].
  
--

<br>

However, in many cases one is .hi[not able] to count all possible outcomes from an experiment.

--

  - For instance, how much money would you like to make five years from now?
  
--

<br>

This answer is probably best given through an .hi-blue[interval], and not an exact amount.

--

That is where .hi-green[continuous random variables] come in.


---

layout: false
class: inverse, middle

# Continuous random variables

---

# Continuous random variables

<br>

A .hi[continuous random variable] can take on .hi-blue[any] real value in an interval.

--

Going back to the salary example, there is an .hi-slate[infinite] number of possible values one can think of.

  - This way, .red[*the probability of each individual value is virtually 0*].

--

<br>

Thus, the probability is best given through a .hi[range] of values.


--

Visually, this can be represented through a .hi-blue[histogram].


---

# Continuous random variables

```{r, dev = "svg", fig.height = 3.5, fig.width = 9}
data <- read_csv("cdc_data.csv")

data %>% 
  ggplot(aes(poverty_rate)) +
  geom_histogram(color = "white", fill = "#800080", binwidth = 1) +
  labs(x = "Poverty rate in the US (1999 - 2017)") +
  easy_x_axis_title_size(13) +
  easy_y_axis_title_size(13)
```


---

# Continuous random variables

Recall that, in a .hi[histogram], we count the number of occurrences of a range values of a variable in a predetermined .hi-blue[interval].

--

These intervals configure the histogram's .red[*bin size*] (or bin width).

--

By dividing the number of counts within each bin by the sample size, we obtain the .hi-blue[relative frequencies] of each range of values.

--

```{r}
data %>% nrow()
```


--

For instance, there are 129 values that fall within the [10 &#8212;11) interval, and 90 values within the [13 &#8212;14) interval.

--

  - .small[129/969 = 0.13 is the .hi[probability] that a randomly chosen poverty rate will lie .red[*between*] 10% and 10.99%.]
  - .small[90/969 = 0.092 is the .hi[probability] that a randomly chosen poverty rate will lie .red[*between*] 13% and 13.99%.]



---

# Continuous random variables

```{r, dev = "svg", fig.height = 4, fig.width = 9}
data %>% 
  ggplot(aes(poverty_rate)) +
  geom_histogram(aes(y = ..density..), color = "white", fill = "#800080", binwidth = 1) +
  labs(x = "Poverty rate in the US (1999 - 2017)") +
  easy_x_axis_title_size(13) +
  easy_y_axis_title_size(13)
```

---

# Continuous random variables

The .hi-slate[sum] of all relative frequencies must add up to 1.

--

In case we draw the histogram with a large number of .hi[small] bins, it is possible to .hi-blue[smooth] its edges.

--

This generates a .hi[density curve].

--

It is possible to approximate this density curve through calculus, obtaining a function *f(x)*.

--

*f(x)* is called a .hi-blue[probability density function].


---

# Continuous random variables

```{r, dev = "svg", fig.height = 3.5, fig.width = 9}
data %>% 
  ggplot(aes(poverty_rate)) +
  geom_histogram(aes(y = ..density..), color = "white", fill = "#800080", binwidth = 0.1, alpha = 0.5) +
  geom_density(alpha = .3, size = 1) +
  labs(x = "Poverty rate in the US (1999 - 2017)") +
  easy_x_axis_title_size(13) +
  easy_y_axis_title_size(13)
```

---

# Continuous random variables

```{r, dev = "svg", fig.height = 3.5, fig.width = 9}
data %>% 
  ggplot(aes(poverty_rate)) +
  geom_histogram(aes(y = ..density..), color = "white", fill = "#800080", binwidth = 1, alpha = 0.5) +
  geom_density(alpha = .3, size = 1) +
  labs(x = "Poverty rate in the US (1999 - 2017)") +
  easy_x_axis_title_size(13) +
  easy_y_axis_title_size(13)
```


---

# Continuous random variables

A *probability density function* (.hi-blue[PDF]) whose range is *a &le; x &le; b* must fulfill the following two .hi-slate[requirements]:

  1. *f(x)* &ge; 0 for all *x* between *a* and *b*;
  
  2. The total *area* under the curve between a and b is 1.
  
--

<br>

Just as with discrete RVs, some random variables show such *specific behaviors* that they can be put into certain categories of PDFs.

--

We will study .hi[two] of the most popular .red[*continuous probability distributions*]:

  - The .hi-blue[Uniform];
  
  - And the .hi-blue[Normal] (Gaussian) distribution.



---

layout: false
class: inverse, middle

# The Uniform distribution

---

# The Uniform distribution

The .hi-blue[Uniform] (*aka* rectangular) distribution is useful when a random variable is .hi[uniformly] or .hi[equally likely] to take on *any value* in a given range.

--

```{r, echo=FALSE, fig.height = 4.5, fig.width = 9, dev = "svg"}
curve(dunif(x, min = 0, max = 10),
      main = "A uniformy distributed RV",
      ylab = "Density")
```





---

# The Uniform distribution

<br><br>

Its .hi-blue[probability density function (PDF)] is given by:

$$
\begin{aligned}
f(x) = \dfrac{1}{b-a}
\end{aligned}
$$


where *a &le; x &le; b*.

---


layout: false
class: inverse, middle

# The Normal distribution


---

# The Normal distribution


The .hi-blue[Normal] (*aka* Gaussian) distribution is the .hi[most popular] probability distribution in Statistics.

--

It is called "Normal" due to its patterns being so .red[*commonly observed*] in data.

--

```{r, echo=FALSE, fig.height = 4.5, fig.width = 9, dev = "svg"}
curve(dnorm(x, mean = 0, sd = 1),
      main = "A normally distributed RV",
      ylab = "Density",
      xlim = c(-3.5, 3.5))
```

---

# The Normal distribution

<br><br>

Its .hi-blue[probability density function (PDF)] is given by:



$$
\begin{aligned}
f(x) = \dfrac{1}{\sigma \sqrt{2\pi}}e^{\frac{1}{2}(\frac{x-\mu}{\sigma})^2} \ ; \ -\infty <x< \infty
\end{aligned}
$$

--

<br>

As it is possible to see from the formula, the Normal distribution is described by .hi[two parameters]:

  1. The population .hi-blue[mean], &mu;;
  
  2. And the population .hi-blue[standard deviation], &sigma;.
  

---

# The Normal distribution

```{r, echo=FALSE, fig.height = 5.5, fig.width = 9, dev = "svg"}
curve(dnorm(x, mean = 0, sd = 1), xlim = c(-5, 5), col = "#1478a7", lwd = 2,
      ylab = "Density",
      main = "Different means, same SD")
curve(dnorm(x, mean = 1, sd = 1), add = TRUE, col = "#701718", lwd = 2)
curve(dnorm(x, mean = -1, sd = 1), add = TRUE, col = '#809b94', lwd = 2)

```

---

# The Normal distribution

```{r, echo=FALSE, fig.height = 5.5, fig.width = 9, dev = "svg"}
curve(dnorm(x, mean = 0, sd = 1), xlim = c(-5, 5), col = "#1478a7", lwd = 2,
      ylab = "Density",
      main = "Same means, different SDs")
curve(dnorm(x, mean = 0, sd = 2), add = TRUE, col = "#701718", lwd = 2)
curve(dnorm(x, mean = 0, sd = 3), add = TRUE, col = '#809b94', lwd = 2)

```



---

layout: false
class: inverse, middle

# Next time: Dealing with continuous distributions


---
exclude: true